---
title: "Assignment: Getting and Cleaning Data Course Project"
output: html_notebook
---

Below is what this script does: 

1. Merges the training and the test sets to create one data set.
2. Extracts only the measurements on the mean and standard deviation for each measurement.
3. Uses descriptive activity names to name the activities in the data set.
4. Appropriately labels the data set with descriptive variable names.
5. From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.

Load packages data.table and reshape2.

```{r}
if (!require("pacman")) {
    install.packages("pacman")
}

pacman::p_load(data.table, reshape2, gsubfn)
```

Get data from zip file.

```{r}
path <- getwd()
url <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
download.file(url, file.path(path, "data.zip"))
unzip(zipfile = "data.zip")
file.remove("data.zip") 
```

Load activity_labels and features.

```{r}
activityLabels <- fread(file.path(path, "UCI HAR Dataset/activity_labels.txt"),
                        col.names = c("classLabels", "activityNames"))

head(activityLabels)
```

```{r}
features <-fread(file.path(path, "/UCI HAR Dataset/features.txt"),
                 col.names = c("index", "featureNames"))

head(features)
```

Extracting mean and std from features.

```{r}
featuresNeeded <- grep("(mean|std)\\(\\)", features[, featureNames])
measurements <- features[featuresNeeded, featureNames]
measurements <- gsubfn( "(^t|^f|Acc|Gyro|Mag|BodyBody|\\(\\))",
                        list(   "t" = "Time",
                                "f" = "Frequency",
                                "Acc" = "Accelerometer",
                                "Gyro" = "Gyroscope",
                                "Mag" = "Magnitude",
                                "BodyBody" = "Body",
                                "()" = ""),
                        measurements)

head(measurements)
```

Load train data. Read in train, filtering based on features needed, using with=False to retain data.frame class.

```{r}
train <- fread(file.path(path, "/UCI HAR Dataset/train/X_train.txt"))[, featuresNeeded, with = FALSE]
setnames(train, colnames(train), measurements) # change column name based on measurement

activityTrain <- fread( file.path(path, "/UCI HAR Dataset/train/y_train.txt"),
                        col.names = "Activity")
subjectTrain <- fread(  file.path(path, "/UCI HAR Dataset/train/subject_train.txt"),
                        col.names = "SubjectNo.")

train <- cbind(activityTrain, subjectTrain, train) # bind all columns together

head(train)
```

Load test data.

```{r}
test <- fread(file.path(path, "/UCI HAR Dataset/test/X_test.txt"))[, featuresNeeded, with = FALSE]
setnames(test, colnames(test), measurements)

activityTest <- fread(  file.path(path, "/UCI HAR Dataset/test/y_test.txt"),
                        col.names = "Activity")
subjectTest <- fread(   file.path(path, "/UCI HAR Dataset/test/subject_test.txt"),
                        col.names = "SubjectNo.")

test <- cbind(activityTest, subjectTest, test) 

head(test)
```

Merge test and train by rows.

```{r}
testTrain <- rbind(train, test)

head(testTrain)
```

Factor Activity column based on activity labels. Use factor() to set own levels and labels.

```{r}
testTrain[["Activity"]] <- factor(testTrain[, Activity], 
                                  levels = activityLabels[["classLabels"]], 
                                  labels = activityLabels[["activityNames"]]) 

head(testTrain)
```

as.factor() to create turn subject numbers into factors.

```{r}
testTrain[["SubjectNo."]] <- as.factor(testTrain[, SubjectNo.])

head(testTrain)
```

Melt then cast the data table.

```{r}
testTrain <- melt.data.table(testTrain, id=c("SubjectNo.", "Activity")) # melt down to variable & value
testTrain <- dcast(testTrain, SubjectNo. + Activity ~ variable, mean) # average of SubjectNo & Activity

head(testTrain)
```

Write final tidy data into new file.

```{r}
fwrite(testTrain, file="tidyData.txt")
```

```{bash}
rm ./tidyData.txt
rm -rf ./UCI\ HAR\ Dataset
```