---
title: "Assignment: Regression Models Course Project"
output: html_notebook
---

You work for Motor Trend, a magazine about the automobile industry. Looking at a data set of a collection of cars, they are interested in exploring the relationship between a set of variables and miles per gallon (MPG) (outcome). They are particularly interested in the following two questions:

* “Is an automatic or manual transmission better for MPG”
* "Quantify the MPG difference between automatic and manual transmissions"

# Exploratory Data Analysis

```{r}
head(mtcars)
```

```{r}
library(dplyr)
mtcars <- mtcars %>%
mutate(am = factor(am, labels = c("automatic", "manual"))) %>%
mutate(vs = factor(vs))
```

## Two samples t test

```{r}
t.test(mtcars$mpg ~ mtcars$am)
```

# Regression Model and Hypothesis testing

## Simple Linear Regression Model

```{r}
fit_am <- lm(mpg ~ am, mtcars)
summary(fit_am) 
```

## Multivariate Linear Regression Model

```{r}
full.model <- lm(mpg ~ ., data = mtcars)
summary(full.model) 
```

```{r}
library(car)
rbind(vif(full.model))
```

## Stepwise regression model

```{r}
bestModel <- step(full.model, direction = "both", trace = FALSE)
summary(bestModel)
```

# Regression diagnostics

```{r}
vif <- cbind(vif(bestModel))
colnames(vif) <- "VIF of bestmodel"
vif
```

## Anova test on nested models

```{r}
fit0 <- lm(mpg ~ am, mtcars)
fit1 <- update(fit0, mpg ~ am + wt)
fit2 <- update(fit1, mpg ~ am + wt + qsec)
fit3 <- update(fit2, mpg ~ am + wt + qsec + disp)
fit4 <- update(fit3, mpg ~ am + wt + qsec + disp + hp)

anova(fit0, fit1, fit2, fit3, fit4)
```
