---
title: "Assignment: Statistical Inference Course Project"
output: html_notebook
---

# Part 1: Simulation Exercise

In this project you will investigate the exponential distribution in R and compare it with the Central Limit Theorem. The exponential distribution can be simulated in R with rexp(n, lambda) where lambda is the rate parameter. The mean of exponential distribution is 1/lambda and the standard deviation is also 1/lambda. Set lambda = 0.2 for all of the simulations. You will investigate the distribution of averages of 40 exponentials. Note that you will need to do a thousand simulations.

Illustrate via simulation and associated explanatory text the properties of the distribution of the mean of 40 exponentials. You should

* Show the sample mean and compare it to the theoretical mean of the distribution.
* Show how variable the sample is (via variance) and compare it to the theoretical variance of the distribution.
* Show that the distribution is approximately normal. Focus on the difference between the distribution of a large collection of random exponentials and the distribution of a large collection of averages of 40 exponentials.

# Part 2: Basic Inferential Data Analysis Instructions

Now in the second portion of the project, we're going to analyze the ToothGrowth data in the R datasets package.

* Load the ToothGrowth data and perform some basic exploratory data analyses
* Provide a basic summary of the data.
* Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose. (Only use the techniques from class, even if there's other approaches worth considering)
* State your conclusions and the assumptions needed for your conclusions.

# Part 1

## Simulation

### Sample exponential distribution

```{r}
set.seed(2021) # for reproducability
nosim <- 1000 # no of simulations
n <- 40 # no of exponentials
lambda <- 0.2 # rate parameter 

simdata <- matrix(rexp(nosim * n, rate=lambda), nosim)
sim_mean <- rowMeans(simdata) # row means

# Calculate mean, sd and variance of sample exp dist
simdata_mean <- mean(sim_mean)
simdata_sd <- sd(sim_mean)
simdata_var <- var(sim_mean)
```

### Theoretical exponential distribution

```{r}
# calculate mean, df and variance of theoretical exp dist
t_mean = 1/lambda
t_sd = (1/lambda) * (1/sqrt(n))
t_var = t_sd^2
```

## Histogram of sample exponential distribution vs Averages of simulated exponentials.

```{r}
par(mfrow = c(1, 2))
hist(simdata,
     col = "steelblue",
     main = "Simulated exponential distribution",
     xlab = "40 random exponentials")
hist(sim_mean,
     col = "red3",
     main = "Averages of Simulated Exponentials",
     xlab = "Average of 40 exponentials")
abline(v = t_mean, col = "green", lwd = 2) # theoretical mean
```

## Comparison between sample and theoretical statistics

```{r}
Sample_stats <- c(simdata_mean, simdata_sd, simdata_var)
Theoretical_stats <- c(t_mean, t_sd, t_var)
diff <- c(abs(t_mean - simdata_mean), abs(t_sd - simdata_sd), t_var - simdata_var)
names <- c("Mean", "Std", "Variance")
data.frame(Sample_stats, Theoretical_stats, diff, row.names =  c("Mean", "Std", "Variance"))
```

## Distribution

### Histogram and Density plot

```{r}
library(ggplot2)
simdata_mean <- data.frame(sim_mean)
ggplot(simdata_mean, aes(sim_mean)) + geom_histogram(binwidth = .3,
                                                    fill = "steelblue",
                                                    color = "black",
                                                    aes(y = ..density..)) +
    geom_density(color = "blue", lwd = 1) + labs(title = "Distribution of Random Exponential Values with 1000 simulations", x = "Average of 40 Exponentials", y = "Density") + stat_function( fun = dnorm,
                                                                                args = list(mean = t_mean, sd = t_sd),
                                                                                color = "red",
                                                                                lwd = 1 ) + theme_bw()
```

### Q-Q plot

```{r}
qqnorm(sim_mean, col="black") # sample distribution
qqline(sim_mean, col="red", lwd=3) #theoretical
```

# Part 2

## Load data

```{r}
data("ToothGrowth")
dt <- ToothGrowth
```

## Basic exploratory Data Analyses

```{r}
str(dt)
```

```{r}
unique(dt$dose)
```

```{r}
dt$dose <- factor(dt$dose)
str(dt)
```

```{r}
summary(dt)
```

```{r}
library(ggplot2)
ggplot(dt, aes(x = dose, y = len, fill = supp)) + geom_boxplot() + ggtitle("Length of tooth growth based on supplement type and dose") + theme_classic()
```

```{r}
library(dplyr)
library(tidyr)

dt %>%
group_by(supp, dose) %>%
summarize(mean = mean(len), .groups='drop') %>%
spread(supp, mean) %>%
mutate(diff = abs(VC - OJ))
```

## t.test Hypothesis Testing

```{r}
dose_half <- filter(dt, dose == 0.5)
dose_one <- filter(dt, dose == 1)
dose_two <- filter(dt, dose == 2)
```

### t-test for 0.5 mg/day dose

```{r}
t.test(len ~ supp, dose_half)
```

###  t-test for 1 mg/day dose

```{r}
t.test(len ~ supp, dose_one)
```

###  t-test for 2 mg/day dose

```{r}
t.test(len ~ supp, dose_two)
```

```{r}
dose <- c(0.5, 1.0, 2.0)
p_value <- c(0.0064, 0.0010, 0.9639)
conf.int <- c("1.72, 8.78", "2.80, 9.06", "-3.80, 3.64")
decision <- c("Reject null", "Reject null", "Do not reject null")
data.frame(dose, conf.int, p_value, decision)
```
