---
title: "swirl Lesson 1: Lattice Plotting System"
output: html_notebook
---

Lattice_Plotting_System. (Slides for this and other Data Science courses may be found at github https://github.com/DataScienceSpecialization/courses/. If you care to use them, they must be downloaded as a zip file and viewed locally. This lesson corresponds to 04_ExploratoryAnalysis/PlottingLattice.)

In another lesson, we gave you an overview of the three plotting systems in R. In this lesson we'll focus on the lattice plotting system.  As we did with the base plotting system, we'll focus on using lattice  to create graphics on the screen device rather than another graphics device.

The lattice plotting system is completely separate and independent of the base plotting system. It's an add-on package so it has to be explicitly loaded with a call to the R function library. We've done this for you. The R Documentation tells us that lattice "is an implementation of Trellis graphics for R. It is a powerful and elegant high-level data visualization system with an emphasis on multivariate data."

Lattice is implemented using two packages. The first is called, not surprisingly, lattice, and it contains code for producing Trellis graphics. Some of the functions in this package are the higher level functions which you, the user, would call. These include xyplot, bwplot, and levelplot.

If xyplot produces a scatterplot, what kind of plot does bwplot produce?

```{r}
"box and whisker"
```

The second package in the lattice system is grid which contains the low-level  functions upon which the lattice package is built. You, the user, seldom call functions from the grid package directly.

Unlike base plotting,  the lattice system does not have a "two-phase" aspect with separate plotting and annotation. Instead all plotting and annotation is done at once with a single function call.

The lattice system, as the base does, provides several different plotting functions. These include xyplot for creating scatterplots, bwplot for box-and-whiskers plots or boxplots, and  histogram for  histograms. There are several others (stripplot, dotplot, splom and levelplot), which we won't cover here.

Lattice functions generally take a formula for their first argument, usually of the form y ~ x. This indicates that y depends on x, so in a scatterplot y would be plotted on the y-axis and x on the x-axis.

Here's an example of  typical lattice plot call,  xyplot(y ~ x | f * g, data). The f and g represent the optional conditioning variables. The * represents interaction between them. Remember when we said that lattice is good for plotting multivariate data? That's where these conditioning variables come into play.

The second argument is the data frame or list from which the variables in the formula should be looked up.  If no data frame or list is passed, then the parent frame is used. If no other arguments are passed, the default values are used.

Recall the airquality data we've used before. We've loaded it again for you. To remind yourself what it looks like run the R command head with airquality as an argument to see what the data looks like.

```{r}
head(airquality)
```

Now try running xyplot with the formula Ozone~Wind as the first argument and the second argument data set equal to airquality.

```{r}
library(lattice)
xyplot(Ozone ~ Wind, data=airquality)
```

Look vaguely familiar? The dots are blue, instead of black, but lattice labeled the axes for you. You can use some of the same graphical parameters (e.g., pch and col) that you used in the base package in calls to lattice functions.

Now rerun xyplot with the formula Ozone~Wind as the first argument and the second argument data set equal to airquality (use the up arrow to save typing). This time add the arguments col set equal to "red", pch set equal to 8, and main set equal to "Big Apple Data".

```{r}
xyplot(Ozone ~ Wind, data=airquality, pch=8, col="red", main="Big Apple Data")
```

Red snowflakes are cool, right? Now that you’ve seen the basic xyplot() and some of its arguments, you might want to experiment more by yourself when you're done with the lesson to discover what other arguments and colors are available. (If you can't wait to experiment, recall that swirl has play() and nxt() functions. At a command prompt, typing play() allows you to leave swirl temporarily so  you can try different R commands at the console. Typing nxt() when you’re done playing brings you back to swirl and you can resume your lesson.) 

Now you'll see how easy it is to generate a multipanel plot using a single lattice command. 

Run xyplot with the formula Ozone~Wind | as.factor(Month) as the first argument and the second argument data set equal to airquality (use the up arrow to save typing). So far, not much is different, right? Add a third argument, layout, set equal to c(5,1). 

```{r}
xyplot(Ozone ~ Wind | as.factor(Month), data=airquality, layout=c(5,1))
```

Note that the default color and plotting character are back. What did the as.factor(Month) do?

```{r}
"Displayed and labeled each subplot with the month's integer"
```

Since Month is a named column of the airquality dataframe we had to tell R to treat it as a factor. To see how this affects the plot, rerun the xyplot command you just ran, but use Ozone ~ Wind | Month instead of Ozone ~ Wind | as.factor(Month) as the first argument.  

```{r}
xyplot(Ozone ~ Wind | Month, data=airquality, layout=c(5, 1))
```

Not as informative, right? The word Month in each panel really doesn't tell you much if it doesn't identify which month it's plotting. Notice that the actual data is the same between the two plots, though.

Lattice functions behave differently from base graphics functions in one critical way. Recall that base graphics functions plot data directly to the graphics device (e.g., screen, or file such as a PDF file). In contrast, lattice graphics functions return an object of class trellis.

The print methods for lattice functions actually do the work of plotting the data on the graphics device. They return "plot objects" that can be stored (but it’s usually better to just save the code  and data). On the command line, trellis objects are auto-printed so that it appears the function is plotting the data.

To see this, create a variable p which is assigned the output of this simple call to xyplot, xyplot(Ozone~Wind,data=airquality).

```{r}
p <- xyplot(Ozone~Wind, data=airquality)
```

Nothing plotted, right? But the object p is around.

Type p or print(p) now to see it.

```{r}
p
```

Like magic, it appears. Now run the R command names with p as its argument.

```{r}
names(p)
```

We see that the trellis object p has 45 named properties, the first of which is "formula" which isn't too surprising. A lot of these properties are probably NULL in value. We've done some behind-the-scenes work for you and created two vectors. The first, mynames, is a character vector of the names in p. The second is a boolean vector, myfull,  which has TRUE values for nonnull entries of p. Run mynames[myfull] to see which entries of p are not NULL.

```{r}
mytrellis <- xyplot(Ozone ~ Wind, data=airquality)
mynames <- names(mytrellis)
myfull <- sapply(mynames, function(nn) !is.null(mytrellis[[nn]]))
```


```{r}
mynames[myfull]
```

Wow! 29 nonNull values for one little plot. Note that a lot of them are like the ones we saw in the base plotting system. Let's look at the values of some of them. Type p[["formula"]] now.
  
```{r}
p[["formula"]]
```

Not surprising, is it? It's a familiar formula. Now look at p's x.limits. Remember the double square brackets and quotes. 

```{r}
p[["x.limits"]]
```

They match the plot, right? The x values are indeed between .37 and 22.03. 

Again, not surprising. Before we wrap up, let's talk about lattice's panel functions which control what happens inside each panel of the plot. The ease of making multi-panel plots makes lattice very appealing. The lattice package comes with default panel functions, but you can customize what happens in each panel. 

Panel functions receive the x and y coordinates of the data points in their panel (along with any optional arguments). To see this, we've created some data for you - two 100-long vectors, x and y. For its first 50 values y is a function of x, for the last 50 values, y is random. We've also defined a 100-long factor vector f which distinguishes between the first and last 50 elements of the two vectors. Run the R command table with f as it argument.

```{r}
set.seed(10)
x <- rnorm(100)
f <- rep(0:1, each=50)
y <- x + f - f * x+ rnorm(100, sd=0.5)
f <- factor(f, labels=c("Group 1", "Group 2"))
```

```{r}
table(f)
```

The first 50 entries of f are "Group 1" and the last 50 are "Group 2". Run xyplot with two arguments. The first is the formula y~x|f, and the second is layout set equal to c(2,1). Note that we're not providing an explicit data argument, so xyplot will look in the environment and see the x and y that we've generated for you.

```{r}
xyplot(y ~ x | f, layout=c(2, 1))
```

To understand this a little better look at the variable v1 we've created for you.

```{r}
rx1 <- range(x[1:50])
rx2 <- range(x[51:100])
ry1 <- range(y[1:50])
ry2 <- range(y[51:100])
v1 <- c(rx1,ry1)
v2 <- c(rx2,ry2)
```


```{r}
v1
```

The first two numbers are the range of the x values of Group 1 and the last two numbers are the range of y values of Group 1. See how they match the values of the left panel (Group 1) in the plot. Now look at v2 which holds the comparable numbers for Group 2.

```{r}
v2
```

Again, the values match the plot. That's reassuring. We've copied some code from the slides for you. To see it, type myedit("plot1.R"). This will open your editor and display the R code in it.

```{r}
p <- xyplot(y ~ x | f, panel = function(x, y, ...) {
    panel.xyplot(x, y, ...)  ## First call the default panel function for 'xyplot'
    panel.abline(h=median(y), lty = 2)  ## Add a horizontal line at the median
})
print(p)
invisible()
```

How many calls to basic lattice plotting functions are there in plot1.R?

```{r}
"1"
```

Note the panel function. How many formal arguments does it have?

```{r}
"3"
```

The panel function has 3 arguments, x, y and ... . This last stands for all other arguments (such as graphical parameters) you might want to include. There are 2 lines in the panel function. Each invokes a panel method, the first to plot the data in each panel (panel.xyplot), the second to draw a horizontal line in each panel (panel.abline). Note the similarity of this last call to that of the base plotting function of the same name.

See how the lines appear. The plot shows two panels because...?

```{r}
"f contains 2 factors"
```

We've copied another piece of similar code, i.e., a call to xyplot with a custom panel function, from the slides. To see it, type myedit("plot2.R"). This will open your editor and display the R code in it.

```{r}
p2 <- xyplot(y ~ x | f, panel=function(x, y, ...) {
    panel.xyplot(x, y, ...)  ## First call default panel function
    panel.lmline(x, y, col=2)  ## Overlay a simple linear regression line
})
print(p2)
invisible()
```

You can see how plot2.R differs from plot1.R, right? 

The regression lines are red because ...?

```{r}
"the custom panel function specified a col argument"
```

Before we close we'll look at how easily lattice can handle a plot with a great many panels. (The sky's the limit.) We've loaded some diamond data for you. It comes with the ggplot2 package. We'll use it just to show off lattice's panel plotting capability.

The data is in the data frame diamonds. Use the R command str to see what it looks like.

```{r}
library(ggplot2)
str(diamonds)
```

So the data frame contains 10 pieces of information for each of 53940 diamonds. Run the R command table with diamonds$color as an argument.

```{r}
table(diamonds$color)
```

We see 7 colors each represented by a letter. Now  run the R command table with two arguments, diamonds$color and diamonds$cut.

```{r}
table(diamonds$color, diamonds$cut)
```

We see a 7 by 5 array with counts indicating how many diamonds in the data frame have a particular color and cut. From the table, which is the most frequent combination?

```{r}
"Ideal cut of color G "
```

To save you some trouble we've defined three character strings for you, labels for the x- and y-axes and a main title. They're in the file myLabels.R, so run myedit on this file to see  them. Remember to put the file name in quotes when you call myedit.

```{r}
myxlab <- "Carat"
myylab <- "Price"
mymain <- "Diamonds are Sparkly!"
```

Now call xyplot with the formula price~carat | color*cut  and data set equal to diamonds. In addition, set the argument strip equal to  FALSE,  pch set equal to 20, xlab to myxlab, ylab to myylab, and main to mymain. The plot may take longer than previous plots because it is bigger.

```{r}
xyplot(price~carat|color*cut, data=diamonds, strip=FALSE, pch=20, xlab=myxlab, ylab=myylab, main=mymain)
```

Pretty cool, right? 35 panels, one for each combination of color and cut. The dots (pch=20) show how prices for the diamonds in each category (panel) vary depending on carat.

Are colors defining the rows or columns of the plot?

```{r}
"columns"
```

Were you curious about that argument strip? I know I was. Now  rerun the xyplot command you just ran (use the up arrow key to retrieve it), this time without the strip argument. 

```{r}
xyplot(price~carat|color*cut, data=diamonds, pch=20, xlab=myxlab, ylab=myylab, main=mymain)
```

The plot shows that the strip argument ....

```{r}
"columns"
```

True or False? Lattice plots are constructed by a series of calls to core functions.

```{r}
"False"
```

True or False? Lattice plots are constructed with a single function call to a core lattice function (e.g. xyplot)

```{r}
"True"
```

True or False? Aspects like margins and spacing are automatically handled and defaults are usually sufficient.

```{r}
"True"
```

True or False? The lattice system is ideal for creating conditioning plots where you examine the same kind of plot under many different conditions.

```{r}
"True"
```

True or False? The lattice system, like the base plotting system, returns a trellis plot object.

```{r}
"False"
```

True or False?  Panel functions can NEVER be customized to modify what is plotted in each of the plot panels.

```{r}
"False"
```

True or False?  Lattice plots can display at most 20 panels in a single plot.

```{r}
"False"
```